Dynamic Modeling, Parameter Estimation and Uncertainty Analysis in \proglang{R}

### 2016-10-13 article.tex
Added Extensions

\pkg{dMod} makes use of derivatives wherever possible. It is one of the core functionalities of the \pkg{cOde} package upon which \pkg{dMod} is based to augment a system of ODEs by its sensitivity equations. If many parameters are involved and the system consists of many states, the number of equations quickly grows. Solving the equations can be considerably accelerated by utilizing compiled code. \pkg{dMod} provides the \code{odemodel} class. An object of class \code{odemodel} is generated from an \code{eqnlist} or \code{eqnvec}. In the background \proglang{C} code for the ODE and the combined system of ODE and sensitivity equations is generated, written to the working directory and compiled. The \code{odemodel} object keeps track of the shared objects. It is the basis of a prediction function $x(t, p)$ generated by the \code{Xs()} command.

@@

\subsection{Parameter uncertainty and identifiability}

@@

\subsection{Steady-state constraints and implicit transformations}

@@

Next, we want to implement the steady-state constraint by an \textit{implicit} parameter transformation, as opposed to the explicit transformation shown above. Let $\dot x = f(x, p)$ be our dynamic system. Then, under certain conditions, we find a function $g(p)$ such that $f\big( g(p), p\big) = 0$ for all $p$, i.e.~$x_S = g(p)$ is a steady state of $f$. The parameter transformation we want to generate is the function $p \mapsto \big(p, g(p)\big)$. Here, the set of outer parameters is the set of the reaction rates  $p$ whereas the set of inner parameters contains these rates and the corresponding steady states as initial value parameters. The root of $f$ must be determined numerically to which end \code{multiroot()} from the \pkg{rootSolve} package is used.

The following code is a reimplementation of the example above. The condition "open" is special in that the reflux rate is modified compared to the reflux rate by which the steady state is computed. This is solved by an event, flipping a switch variable from 0 to 1 and thereby replacing the rate \code{reflux} by \code{reflux_open}.

@@

>>>>

\pkg{dMod} makes use of derivatives wherever possible. It is one of the core functionalities of the \pkg{cOde} package upon which \pkg{dMod} is based to augment a system of ODEs by its sensitivity equations. If many parameters are involved and the system consists of many states, the number of equations quickly grows. Solving the equations can be considerably accelerated by utilizing compiled code. \pkg{dMod} provides the \code{odemodel} class. An object of class \code{odemodel} is generated from an \code{eqnlist} or \code{eqnvec}. In the background, \proglang{C} code for the ODE and the combined system of ODE and sensitivity equations is generated, written to the working directory and compiled. The \code{odemodel} object keeps track of the shared objects. It is the basis of a prediction function $x(t, p)$ generated by the \code{Xs()} command.

@@

\subsection{Parameter uncertainty and identifiability}
\label{chap_uncertainty}

@@

\subsection{Steady-state constraints and implicit transformations}
\label{chap_steady_state}

@@

Alternatively, we want to implement the steady-state constraint by an \textit{implicit} parameter transformation, as opposed to the explicit transformation shown above. Let $\dot x = f(x, p)$ be our dynamic system. Then, under certain conditions, we find a function $g(p)$ such that $f\big( g(p), p\big) = 0$ for all $p$, i.e.~$x_S = g(p)$ is a steady state of $f$. The parameter transformation we want to generate is the function $p \mapsto \big(p, g(p)\big)$. Here, the set of outer parameters is the set of the reaction rates  $p$ whereas the set of inner parameters contains these rates and the corresponding steady states as initial value parameters. The root of $f$ must be determined numerically to which end \code{multiroot()} from the \pkg{rootSolve} package is used.

The following code is a reimplementation of the example above. The condition "open" is special in the sense that the reflux rate is modified compared to the reflux rate by which the steady state is computed. This is solved by an event, flipping a switch variable from 0 to 1 and thereby replacing the rate \code{reflux} by \code{reflux_open}.

@@

\subsection{Lie-group symmetry detection}
In Section \ref{chap_uncertainty}, profile likelihood computation showed the existence of both practically and structurally non-identifiable parameters. While practical non-identifiabilities might only be removed by additional experiments, structural non-identifiabilties are connected to Lie-group symmetries, i.e. parameter transformations \[\Psi: (x,\theta) \Longrightarrow (x^*, \theta^*)\] that preserve model predictions of the observables: \[g(x^*, \theta^*) = g(x,\theta)\,.\] Based on \cite{merkt2015higher}, the \code{symmetryDetection()} command outputs a list of available symmetry transformations separated into scaling and translational invariances. For example, the code
\begin{CodeChunk}
\begin{CodeInput}
reactions <- NULL
reactions <- addReaction(reactions, "TCA_buffer", "TCA_cell",
                         rate = "import_buffer*TCA_buffer",
                         description = "Uptake")
reactions <- addReaction(reactions, "TCA_cell", "TCA_buffer",
                         rate = "exp_sinus*TCA_cell",
                         description = "Sinusoidal export")
reactions <- addReaction(reactions, "TCA_cell", "TCA_cana",
                         rate = "exp_cana*TCA_cell",
                         description = "Canalicular export")
reactions <- addReaction(reactions, "TCA_cana", "TCA_buffer",
                         rate = "reflux*TCA_cana",
                         description = "Reflux into the buffer")

observables <- eqnvec(
  buffer = "s*TCA_buffer",
  cellular = "s*(TCA_cana + TCA_cell)"
)
symmetryDetection(as.eqnvec(reactions), observables)
\end{CodeInput}
\end{CodeChunk}
outputs one scaling invariance for the parameters \code{s}, \code{TCA_buffer}, \code{TCA_cana} and \code{TCA_cell} which is in agreement with the identifiability analysis by the method of profile likelihood. Here, the scaling invariance could be removed by fixing one of the parameters to a certain value unequal to zero\footnote{Fixing $\mbox{TCA\_buffer}=0$ does not remove the invariance, since the corresponding symmetry transformation is not valid (\cite{merkt2015higher}).}. For the case of a translational invariance all values were allowed.

\subsection{Analytical steady-state constraints}
While for the the present model, the explicit steady-state implementation could be easily calculated by hand, this might be much more challenging for model classes with high amounts of states and parameters. For many of these cases, the \code{steadyStates()} command outputs an analytical steady-state solution that can be incorporated in the model as an additional parameter transformation. Based on \cite{rosenblatt2016customized}, the steady-state constraint is solved for a combination of state variables and kinetic parameters while positivity of the solution is ensured. As the paper states, the approach outperforms common methods of steady-state implementation with respect to reliability and performance of the optimization process. The corresponding code reads:
\begin{CodeChunk}
\begin{CodeInput}
SS_equations <- steadyStates(reactions)
pSS <- pSS + P(SS_equations)
\end{CodeInput}
\end{CodeChunk}
In addition, \code{steadyStates()} prints conserved quantities that have been found during the calculation. In order to keep the steady-state solution as simple as possible, one of the state variables that contribute to the conserved quantity are left unconstrained.
